Staged Abstract Interpreters

### 2019-08-15 paper/40_absinterp.tex
typos

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We first present a
lattice framework with binding-time abstraction, as well as simple abstract
domains such as pwoer set lattice.
Then we show the abstract components, namely abstract values, abstract
environments, and abstract stores. The abstract interpreter we construct in this
section is similar to \citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For
simplicity, it is also context/path/flow-insensitive by our choice. In Section
\ref{cfa}, we will instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

We first widen the concrete values to @AbsValue@: numbers are lifted to a
singleton object @IntTop@ representing the set of all numbers; closures remain
the same. Then the type @Value@ is a set of @AbsValue@ (we abuse the type names
a bit, as the generic interface already defined the abstract type @Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After defining the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
values occurred at runtime.

\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

To prevent the abstract interpreter from falling into non-termination when
computing the fixed-point, the ADI approach uses a co-inductive mechanism
consisting of two caches that remember the input and output of functions.
Here we first provide the necessary type definitions, and later we will describe
the algorithm in detail. A @Cache@ is a mapping from configurations @Config@ to
sets of value-store pairs, where the configuration is a triple of current
expression being evaluated, environment and store. Intuitively, a cache memoizes
the corresponding values and stores for a given program configuration.

\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
\end{lstlisting}

Now, the store is a mapping from addresses to sets of abstract values. This can
be justied by the approximation nature of abstract interpretation and
nondeterminism in the abstract interpreter.
First, when analyzing a conditional experssion, we may not have enough
information on which branch will be taken, thus a sound treatment is to
nondeterministically explore both branches, i.e., compute the join of two path.
Second, when applying a function, for instance, @f(a)@, the possible lambda
terms of @f@ is computed approximately. As a result, we may retrieve a set of
closures when dereferencing the address of @f@ and need to apply all of them.

\paragraph{Monads for Abstract Interpretation} Compared with the concrete
interpreter that uses reader and state effects, the abstract interpreter
further introduces a nondeterminism effect and another reader and state effect
for the two caches. The nondeterminism effect is represented by the
@Set[M[_], A]@ monad transformer, where @M@ is the inner monad being
transformed, and @A@ is the type of elements in the set. We use a @ReaderT@ for
one cache that will not be changed during one fixed-point iteration, and use a
@StateT@ for another cache that will be constantly updated during analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permuatation of the
moand stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):

\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

Similar to the concrete scenario, we sketch the implementation of @flatMap@ for
@SetT@ transformer and omit the rest. The value @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. The @flatMap@ method takes a function that transforms each
element of type @A@ from @M[Set[A]]@ into a @SetT[M, B]@ value, and then folds
all the transformed values into a single monadic value of type @Set[M, B]@.

\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M,B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

Such monad stack scheme leads to a different type of result type. In the concrete
setting, we only have a reader monad (for environment) and a state monad (for
store), and the reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. In the abstract setting, we have a @SetT@ inside of the
environment monad and store monad, which becomes the container type of the pair
of values and stores, i.e., @Set[(Value, Store)]@. Also note that the reader and
state monad for caches live internally of the nondeterminism monad, therefore
the cache type of state monad becomes adjacent to the set type.

\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed to adapt
the new monad stack scheme and value domains. A notable change is that the store
update operation joins the value with the existing values, for the same address.
As what mentioned before, we may have to explore both branches for conditionals:
in @br0@ method, we combine the results using @mplus@ from @MonadPlus@, which
requires the value domains are join-semilattices (in our case @Set@s and
@Map@s).

\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

The value representation of lambda terms is still a defunctionalized closure
@CloV@, but we lift it to a singleton set domain.

\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is now the first argument @rator@ is a set of closures (remember
that @Value@ is now an alias of @Set[AbsValue]@), so we can not directly
destruct it by pattern matching. Instead, we use an auxiliary function @lift_nd@
that takes a set and lifts the elements into the monad stack. Then we can still
implement @ap_clo@ in monadic style, where the closures come from the @SetT@ and
thus the nondeterminism can be naturally handled. The light gray code shows what
is added from the concrete @ap_clo@.

\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](rator)}!
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been used in other abstract
interpretater or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation, where the @in@ cache stores the result from last
iteration, and @out@ cache is used for updates in current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In out monad stack, the @in@ and @out@ cache modeled by a reader effect and
state effect, respectively. We first defined several methods to manipulate
the two caches through the monad stack; the implementations are elided.

\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

>>>>

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We first present a
lattice framework with binding-time abstraction, as well as simple abstract
domains such as power set lattice.
Then we show the abstract components, namely abstract values, abstract
environments, and abstract stores. The abstract interpreter we construct in this
section is similar to \citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For
simplicity, it is also context/path/flow-insensitive by our choice. In Section
\ref{cfa}, we will instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

We first widen the concrete values to @AbsValue@: numbers are lifted to a
singleton object @IntTop@ representing the set of all numbers; closures remain
the same. Then the type @Value@ is a set of @AbsValue@ (we abuse the type names
a bit, as the generic interface already defined the abstract type @Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After defining the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
values occurred at run-time.
\begin{lstlisting}
  trait AbstractComponents extends Semantics {
    sealed trait AbsValue
    case object IntTop extends AbsValue
    case class CloV(lam: Lam, env: Env) extends AbsValue
    type Value = Set[AbsValue]; case class Addr(x: String)
    ...
  }
\end{lstlisting}

To prevent the abstract interpreter from falling into non-termination when
computing the fixed-point, the ADI approach uses a co-inductive mechanism
consisting of two caches that remember the input and output of functions.
Here we first provide the necessary type definitions, and later we will describe
the algorithm in detail. A @Cache@ is a mapping from configurations @Config@ to
sets of value-store pairs, where the configuration is a triple of current
expression being evaluated, environment and store. Intuitively, a cache memoizes
the corresponding values and stores for a given program configuration.
\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
\end{lstlisting}

Now, the store is a mapping from addresses to sets of abstract values. This can
be justified by the approximation nature of abstract interpretation and
nondeterminism in the abstract interpreter.
First, when analyzing a conditional expression, we may not have enough
information on which branch will be taken, thus a sound treatment is to
nondeterministically explore both branches, i.e., compute the join of two path.
Second, when applying a function, for instance, @f(a)@, the possible lambda
terms of @f@ is computed approximately. As a result, we may retrieve a set of
closures when dereferencing the address of @f@ and need to apply all of them.

\paragraph{Monads for Abstract Interpretation} Compared with the concrete
interpreter that uses reader and state effects, the abstract interpreter
further introduces a nondeterminism effect and another reader and state effect
for the two caches. The nondeterminism effect is represented by the
@Set[M[_], A]@ monad transformer, where @M@ is the inner monad being
transformed, and @A@ is the type of elements in the set. We use a @ReaderT@ for
one cache that will not be changed during one fixed-point iteration, and use a
@StateT@ for another cache that will be constantly updated during analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permutation of the
monad stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):
\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

Similar to the concrete scenario, we sketch the implementation of @flatMap@ for
@SetT@ transformer and omit the rest. The value @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. The @flatMap@ method takes a function that transforms each
element of type @A@ from @M[Set[A]]@ into a @SetT[M, B]@ value, and then folds
all the transformed values into a single monadic value of type @Set[M, B]@.
\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M,B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

Such monad stack scheme leads to a different type of result type. In the concrete
setting, we only have a reader monad (for environment) and a state monad (for
store), and the reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. In the abstract setting, we have a @SetT@ inside of the
environment monad and store monad, which becomes the container type of the pair
of values and stores, i.e., @Set[(Value, Store)]@. Also note that the reader and
state monad for caches live internally of the nondeterminism monad, therefore
the cache type of state monad becomes adjacent to the set type.
\begin{lstlisting}
  type Result = (R[Set[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Primitive Operations} The primitive operations are changed to adapt
the new monad stack scheme and value domains. A notable change is that the store
update operation joins the value with the existing values, for the same address.
As what mentioned before, we may have to explore both branches for conditionals:
in @br0@ method, we combine the results using @mplus@ from @MonadPlus@, which
requires the value domains are join-semilattices (in our case @Set@s and
@Map@s).
\begin{lstlisting}
  def set_store(αv: (Addr, Value)): AnsM[Unit] = liftM(StateTMonad.mod(σ => σ ⊔ Map(αv)))
  def br0(test: Value, thn: => Ans, els: => Ans): Ans = ReaderTMonadPlus.mplus(thn, els)
\end{lstlisting}

The value representation of lambda terms is still a defunctionalized closure
@CloV@, but we lift it to a singleton set domain.
\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is now the first argument @rator@ is a set of closures (remember
that @Value@ is now an alias of @Set[AbsValue]@), so we can not directly
destruct it by pattern matching. Instead, we use an auxiliary function @lift_nd@
that takes a set and lifts the elements into the monad stack. Then we can still
implement @ap_clo@ in monadic style, where the closures come from the @SetT@ and
thus the nondeterminism can be naturally handled. The light gray code shows what
is added from the concrete @ap_clo@.
\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](rator)}!
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been used in other abstract
interpreter or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation, where the @in@ cache stores the result from last
iteration, and @out@ cache is used for updates in current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In out monad stack, the @in@ and @out@ cache modeled by a reader effect and
state effect, respectively. We first defined several methods to manipulate
the two caches through the monad stack; the implementations are elided.
\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

### 2019-08-15 paper/50_stageabs.tex
typos

Soundness is the central concern of static analysis, and as such, is vital
for prospective users. We now briefly examine how our
approach preserves the correctness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here is based on empirical evidence; a rigorous proof of soundness
preservation of the staged abstract interpreters remains an open question.

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistence, i.e., the generated program does not change the observational
    behaviors of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal Form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularly.
  \item We construct the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structures we used here are simply
    black boxes that wrap the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

>>>>

Soundness is the central concern of static analysis, and as such, is vital
for prospective users. We now briefly examine how our
approach preserves the correctness and soundness of the analysis, under the
assumption that the unstaged one is sound. It is worth noting that the rationale
listed here is based on empirical evidence; a rigorous proof for soundness
preservation of the staged abstract interpreters remains an open question.

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistence, i.e., the generated program does not change the observational
    behaviors of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal Form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularly.
  \item We build the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structures we used here are simply
    black boxes that wrap the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

### 2019-08-15 paper/60_cases.tex
typos

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is 0-CFA
for a small CPS language consisting of abstractions, applications, recursion and
primtive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., @compCall@, @compApp@, etc.).
Before refactoring into AC, these functions would take the store as a
regular argument; now every function returns a value of type @CompAnalysis@,
which is a function (closure) value that takes and returns stores. After the
first run of analysis, we have traversed the AST and obtained a single closure
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now we can invoke the compiled closure with an initial
store to complete the analysis.

@@

On the other side, our approach does exactly the same thing through staging: the
AST are static, and stores are dynamic, therefore we annotate the type @Store@
with @Rep@ (shown in Figure \ref{compare_ac_sai} (right)). In fact, the only
changes of our appraoch using LMS framework are the types of @Store@ to
@Rep[Store]@, since they will be known at the next stage; other parts do not need to change. 
The generated code
will just looks-up and update the store. The functions at current stage are
eliminated in the residual program, therefore we do not generate higher-order
functions, which provides additional performance improvement. Nevertheless,
the support from LMS framework is necessary, such as staging support for @Map@,
we consider these efforts are reusable and complete orthogonal to the
implementation of analysis.

@@

We then integrate the timestamp into our moand stack by adding a staged
@StateT@ monad. The result type is passingly get updated to contain a set of
tuples, where the time is added.
\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to next-stage value
\end{lstlisting}

@@

A commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swaping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

The idea of abstract garbage collection is shown in Figure \ref{fixgc}. We modify the @fix@
function with caching (from Section \ref{sai}) to @fix_gc@, but the caching part
is omitted. Several auxiliary functions are used: @ask_root@ retrieves the
current root addresses from the moand stack; @root@ compute the reachable
addresses from a value; and @gc@ cleans up the store according to the reachable
addresses and returns a new store. For the generic interpreter, we also need to
compute the root addresses for compound expressions, readers may refer to
\cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

One of the challenges of modern static analysis is that program
usually depends on large libraries programs
\cite{toman_et_al:LIPIcs:2017:7121}. If we are able to analyze
programs and libraries separately and reuse the result without losing
precision, then we can reduce the overhead of repeatedly analyzing
libraries. Some static analyzers compute a summary for a function or a
module, which can be reused later. However, when lacking the context
information, those analyses can be too conservative or unsound, which
both lead to imprecision. In this section, we show that
meta-programming can provide a trade-off between summary-based modular
analysis and slow whole-program analysis, by ultilizing an existing
whole-program analyzer.

@@

The specialization (0CFA and without store-widening) of @close@ on
@map@ generates the following next-stage Scala code, that is, the
function @x39@:
\begin{lstlisting}
  val x39 = {(x40: List[Set[AbsValue]],          // argument value list
              x41: Map[Addr, Set[AbsValue]],     // store
              x42: Map[Config, Set[ValSt]],      // in cache
              x43: Map[Config, Set[ValSt]]) =>   // out cache
    val x50 = List[Addr](ZCFAAddr("xs"), ZCFAAddr("f")) // addresses of arguments
    val x51 = x50.zip(x40)                              // addresses paired with their values
    val x61 = x51.foldLeft (x41) { case (x52, x53) =>   // prepare a new store that joins
      val x54 = x53._1;  val x55 = x53._2               //   the arguments and their values
        val x58 = x52.getOrElse(x54, x57)               //   into the latest store, intially x41
        val x59 = x58.union(x55)
        x52 + (x54 -> x59)
    } ...
  }
\end{lstlisting}

We borrow the notation traditional modular analysis and realize that
such generated function is a reusable artifact which can be used
modularly.  The generated functions already eliminate the
interpretation overhead and monadic overhead, therefore can be
consider as partial summaries. Although, we still need to run the
generated function at next stage to obtain the analysis result.  At
next stage, we can provide different abstract arugments, store and
caches, depending on the context where it is called, thus the analysis
is modular.

Compared with traditional summary-based analysis, our partial
summaries are represented by executable functions that are
parameterized to the context information. In our approach, the
transition from a whole-program analyzer to modular staged analyzer is
mechanized and for free from staging. We do not need to design any
intermediate representation or data structures of summaries.  In some
scenario, traditional summary-based analyses are able to carry more
information within a mdoule, e.g., the summary of their internal
submodules. The staged approach presented here may still need to run
through the fixed-point for submodules.

\iffalse
\paragraph{Partially-Static Context} Under a polyvariant analysis (like
$k$-CFA), such summary generated by specializing static analyzer even could do
more for calling context by partially-static data optimization. For example,
consider that we have a library consists of three functions @f@, @g@ and @h@,
where @f@ calls @g@ and @h@ internally and returns @y@ fianlly (we extend the
language to have multiple arguments). @g@ and @h@ do not depend on @f@. Now we
would like to modularlly specialize these functions, starting from @f@ with an
initial environment containing @f@, @g@ and @h@.
\begin{lstlisting}
  def f(a, b, c) = val x = g(a, b); val y = h(x, c);  y
\end{lstlisting}

@@

Now we consider in a first-order imperative language, we may care more about the
data-flow because the control-flow is relatively easy to obtain. In this
section, we show the staging of other abstract domains, particularly an interval
domain for numbers. It has been shown that specializing abstract domains with
respect to the structure of analyzed program significantly improves the
performance: a recent example is online decomposition of polyhedra
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143}. In this
section, we first show how to support imperative language features in the
generic abstract interpreter. Then we present a similar idea for the interval
domain and show that the specialization is feasible by staging systematically.

@@

To evaluates a @while@ loop statement, we evaluate the condition first. Then
similar to how we treat @branch0@, we have a generic @branch@ function but works
on boolean values. For the @true@ branch, we recursively call @ev@ on a newly
constructed expression @Seq(e, While(t, e))@ meaning that first evaluates @e@
and the repeat the loop. Otherwise, for the other branch, we simply return a
void value and current store.
\begin{lstlisting}
case While(t, e) =>
  val (tv, t$\sigma$) = ev(t, $\rho$, $\sigma$)
  branch(tv, ev(Seq(e, While(t, e)), $\rho$, t$\sigma$), (void(), t$\sigma$))
\end{lstlisting}

>>>>

\paragraph{Original AC}
The analysis presented by \citeauthor{Boucher:1996:ACN:647473.727587} is 0-CFA
for a small CPS language consisting of abstractions, applications, recursion and
primitive operators. Figure \ref{compare_ac_sai} (left) shows the AC
implementation: different syntactic constructs are handled in different functions
(e.g., @compCall@, @compApp@, etc.).
Before refactoring into AC, these functions would take the store as a
regular argument; now every function returns a value of type @CompAnalysis@,
which is a function (closure) value that takes and returns stores. After the
first run of analysis, we have traversed the AST and obtained a single closure
takes a store and returns a store. Therefore, the ASTs are static terms, and the
stores are dynamic terms. Now we can invoke the compiled closure with an initial
store to complete the analysis.

@@

On the other side, our approach does exactly the same thing through staging: the
AST are static, and stores are dynamic, therefore we annotate the type @Store@
with @Rep@ (shown in Figure \ref{compare_ac_sai} (right)). In fact, the only
changes of our approach using LMS framework are the types of @Store@ to
@Rep[Store]@, since they will be known at the next stage; other parts do not need to change. 
The generated code
will just looks-up and update the store. The functions at current stage are
eliminated in the residual program, therefore we do not generate higher-order
functions, which provides additional performance improvement. Nevertheless,
the support from LMS framework is necessary, such as staging support for @Map@,
we consider these efforts are reusable and complete orthogonal to the
implementation of analysis.

@@

We then integrate the timestamp into our monad stack by adding a staged
@StateT@ monad. The result type is passingly get updated to contain a set of
tuples, where the time is added.
\begin{lstlisting}
  def tick(e: Expr): AnsM[Unit] = for {
    τ <- get_time; u <- liftM(StateTMonad.put((e::τ).take($k$)))
  } yield u
  def alloc(x: String): AnsM[Addr] = for {
    τ <- get_time
  } yield unit(KCFAAddr(x, τ)) // we use unit to turn a current-stage constant to next-stage value
\end{lstlisting}

@@

A commonly used technique to improve the running time in control-flow analysis
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, store-widening analysis can be easily
implemented in monadic abstract interpreter by swapping the @StateT@ (for store)
and the @SetT@ (for nondeterminism). In this section, we show this optimization
is orthogonal to staging and we just need to swap the same \textit{staged}
version of monads. After the adjustment, we have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

The idea of abstract garbage collection is shown in Figure \ref{fixgc}. We modify the @fix@
function with caching (from Section \ref{sai}) to @fix_gc@, but the caching part
is omitted. Several auxiliary functions are used: @ask_root@ retrieves the
current root addresses from the monad stack; @root@ compute the reachable
addresses from a value; and @gc@ cleans up the store according to the reachable
addresses and returns a new store. For the generic interpreter, we also need to
compute the root addresses for compound expressions, readers may refer to
\cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

One of the challenges of modern static analysis is that program
usually depends on large libraries programs
\cite{toman_et_al:LIPIcs:2017:7121}. If we are able to analyze
programs and libraries separately and reuse the result without losing
precision, then we can reduce the overhead of repeatedly analyzing
libraries. Some static analyzers compute a summary for a function or a
module, which can be reused later. However, when lacking the context
information, those analyses can be too conservative or unsound, which
both lead to imprecision. In this section, we show that
meta-programming can provide a trade-off between summary-based modular
analysis and slow whole-program analysis, by utilizing an existing
whole-program analyzer.

@@

The specialization (0CFA and without store-widening) of @close@ on
@map@ generates the following next-stage Scala code, that is, the
function @x39@:
\begin{lstlisting}
  val x39 = {(x40: List[Set[AbsValue]],          // argument value list
              x41: Map[Addr, Set[AbsValue]],     // store
              x42: Map[Config, Set[ValSt]],      // in cache
              x43: Map[Config, Set[ValSt]]) =>   // out cache
    val x50 = List[Addr](ZCFAAddr("xs"), ZCFAAddr("f")) // addresses of arguments
    val x51 = x50.zip(x40)                              // addresses paired with their values
    val x61 = x51.foldLeft (x41) { case (x52, x53) =>   // prepare a new store that joins
      val x54 = x53._1;  val x55 = x53._2               //   the arguments and their values
        val x58 = x52.getOrElse(x54, x57)               //   into the latest store, initially x41
        val x59 = x58.union(x55)
        x52 + (x54 -> x59)
    } ...
  }
\end{lstlisting}

We borrow the notation traditional modular analysis and realize that
such generated function is a reusable artifact which can be used
modularly.  The generated functions already eliminate the
interpretation overhead and monadic overhead, therefore can be
consider as partial summaries. Although, we still need to run the
generated function at next stage to obtain the analysis result.  At
next stage, we can provide different abstract arguments, store and
caches, depending on the context where it is called, thus the analysis
is modular.

Compared with traditional summary-based analysis, our partial
summaries are represented by executable functions that are
parameterized to the context information. In our approach, the
transition from a whole-program analyzer to modular staged analyzer is
mechanized and for free from staging. We do not need to design any
intermediate representation or data structures of summaries.  In some
scenario, traditional summary-based analyses are able to carry more
information within a module, e.g., the summary of their internal
sub-modules. The staged approach presented here may still need to run
through the fixed-point for sub-modules.

\iffalse
\paragraph{Partially-Static Context} Under a polyvariant analysis (like
$k$-CFA), such summary generated by specializing static analyzer even could do
more for calling context by partially-static data optimization. For example,
consider that we have a library consists of three functions @f@, @g@ and @h@,
where @f@ calls @g@ and @h@ internally and returns @y@ finally (we extend the
language to have multiple arguments). @g@ and @h@ do not depend on @f@. Now we
would like to modularly specialize these functions, starting from @f@ with an
initial environment containing @f@, @g@ and @h@.
\begin{lstlisting}
  def f(a, b, c) = val x = g(a, b); val y = h(x, c);  y
\end{lstlisting}

@@

Now we consider in a first-order imperative language, we may care more about the
data-flow because the control-flow is relatively easy to obtain. In this
section, we show the staging of other abstract domains, particularly an interval
domain for numbers. It has been shown that specializing abstract domains with
respect to the structure of analyzed program significantly improves the
performance: a recent example is online decomposition of polyhedron
\cite{DBLP:conf/popl/SinghPV17, Singh:2017:PCD:3177123.3158143}. In this
section, we first show how to support imperative language features in the
generic abstract interpreter. Then we present a similar idea for the interval
domain and show that the specialization is feasible by staging systematically.

@@

To evaluates a @while@ loop statement, we evaluate the condition first. Then
similar to how we treat @branch0@, we have a generic @branch@ function but works
on Boolean values. For the @true@ branch, we recursively call @ev@ on a newly
constructed expression @Seq(e, While(t, e))@ meaning that first evaluates @e@
and the repeat the loop. Otherwise, for the other branch, we simply return a
void value and current store.
\begin{lstlisting}
case While(t, e) =>
  val (tv, t$\sigma$) = ev(t, $\rho$, $\sigma$)
  branch(tv, ev(Seq(e, While(t, e)), $\rho$, t$\sigma$), (void(), t$\sigma$))
\end{lstlisting}

### 2019-08-15 paper/70_eval.tex
revision
typos

We also observe that much overhead of unstaged analyzers are from the monadic
layers. Thus our evaluation result provides an abstraction without regret
approach to construct optimizing abstract interpreters -- the user may write a
high-level modular and composable abstract interpreter, and then derive a staged
version in a principled way that runs significantly faster.

@@

>>>>

@@

Our evaluation result provides an abstraction without regret
approach to construct optimizing abstract interpreters -- the user may write a
high-level modular and compositional abstract interpreter, and then derive a
staged version in a principled way that runs significantly faster.
